### 2015-04-29 stat347/hw2.tex

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
STAT 302 Homework 3 \hfill Aaron Maurer
\vspace{2mm}
\hrule
\vspace{2mm}

\begin{itemize}
    \item[1.]
        \begin{itemize}
            \item[(a)]
                \begin{align*}
                    P(D \st P) &= \frac{P(P\st D)P(D)}{P(P\st D)P(D) +  P(P\st D^c)P(D^c)} \\
                               &= \frac{p_0p_1}{p_0p_1 +  (1-p_0)p_1} \\
                \end{align*}
            \item[(b)]
                Here is my R code:
                \lstinputlisting[firstline=18,lastline=25]{hw3.R}
            \item[(c)/(d)]
                I have produced the results for $a=.5,1 \,\&\, 2$ in one table, each time with confidence intervals based off of $10,000$ iterations.
                \FloatBarrier
                \input{hw3/pr1_cd.tex}
                \FloatBarrier
                We can observe the sensitivity to $a$ by noting that the confidence intervals get pulled towards $.5$ as $a$ increases. The distinction isn't huge though; with $n=20$ we note the bounds change by as much as $.1$, but with $n=80$, the change is at most near $.01$.
            \item[(e)]
                I have simulated the portion of the time the true $\theta$ (th) was below the CI lower bound (LB) and the portion the time it was above the upper bound (UB) for each $a$ and the given $p$. The simulation drew the $X$ $1,000$ times, and for each of these generated a confidence interval based on another $1,000$ random draws based on the given $X$.
                \FloatBarrier
                \input{hw3/pr1_e.tex}
                \FloatBarrier
                In general, $a=.05$ seems to have proper, or very close to proper, frequentist coverage properties. However, in the cases where there are $p$ near $0$ or $1$, this seems to be less true, with the worst case being $p_0=.95, p_1=.95, p_2=0$, where we never see $\theta$ fall below the CI. For different values of $a$ though, we seem to consistently see either higher or lower error rates on each side of the CI.
        \end{itemize}
    \item[2.]
        \begin{itemize}
            \item[i)]
                We can derive $\E(\theta_j)$ as such:
                \begin{align*}
                    \E(\theta_j) &= \idotsint_{\sum_{i=1}^k\theta_i=1}\theta_i \frac{\Gamma(\sum_{i=1}^k\alpha_i)}{\prod_{i=1}^k\Gamma(\alpha_i)} \left(\prod_{i=1}^{k}\theta_i^{\alpha_i-1}\right)d\theta_1...d\theta_k \\
                    \E(\theta_j) &= \frac{\Gamma(1+\alpha_j)\Gamma(\sum_{i=1}^k\alpha_i)}{\Gamma(\alpha_j)\Gamma(1+\sum_{i=1}^k\alpha_i)}   \idotsint_{\sum_{i=1}^k\theta_i=1}\frac{\Gamma(1+\sum_{i=1}^k\alpha_i)}{\Gamma(1+\alpha_j)\prod_{i\neq j}\Gamma(\alpha_i)} \left(\theta_i^{\alpha_i}\prod_{i\neq j}\theta_i^{\alpha_i-1}\right)d\theta_1...d\theta_k \\
                    \E(\theta_j) &= \frac{\Gamma(1+\alpha_j)\Gamma(\sum_{i=1}^k\alpha_i)}{\Gamma(\alpha_j)\Gamma(1+\sum_{i=1}^k\alpha_i)}\\
                    \E(\theta_j) &= \frac{\alpha_j}{\sum_{i=1}^k\alpha_i}\\
                \end{align*}
                Then, as a first step to get the variance,
                \begin{align*}
                    \E(\theta_j^2) &= \idotsint_{\sum_{i=1}^k\theta_i=1} \theta_i^2 \frac{\Gamma(\sum_{i=1}^k\alpha_i)}{\prod_{i=1}^k\Gamma(\alpha_i)} \left(\prod_{i=1}^{k}\theta_i^{\alpha_i-1}\right)d\theta_1...d\theta_k \\
                    \E(\theta_j^2) &= \frac{\Gamma(2+\alpha_j)\Gamma(\sum_{i=1}^k\alpha_i)}{\Gamma(\alpha_j)\Gamma(2+\sum_{i=1}^k\alpha_i)}   \idotsint_{\sum_{i=1}^k\theta_i=1}\frac{\Gamma(2+\sum_{i=1}^k\alpha_i)}{\Gamma(2+\alpha_j)\prod_{i\neq j}\Gamma(\alpha_i)} \left(\theta_i^{1+\alpha_i}\prod_{i\neq j}\theta_i^{\alpha_i-1}\right)d\theta_1...d\theta_k \\
                    \E(\theta_j^2) &= \frac{\Gamma(2+\alpha_j)\Gamma(\sum_{i=1}^k\alpha_i)}{\Gamma(\alpha_j)\Gamma(2+\sum_{i=1}^k\alpha_i)}\\
                    \E(\theta_j^2) &= \frac{\alpha_j(\alpha_j+1)}{\left(\sum_{i=1}^k\alpha_i+1\right)\sum_{i=1}^k\alpha_i}\\
                \end{align*}
                Which we can use to get
                \begin{align*}
                    \var(\theta_j) &= \E(\theta_j^2) - \E(\theta_j)^2 \\
                    \var(\theta_j) &= \frac{\alpha_j^2+\alpha_j}{\left(\sum_{i=1}^k\alpha_i+1\right)\sum_{i=1}^k\alpha_i} - \frac{\alpha_j^2}{\left(\sum_{i=1}^k\alpha_i\right)^2} \\
                    \var(\theta_j) &= \frac{\alpha_j^2\left(\sum_{i=1}^k\alpha_i\right)+\alpha_j\left(\sum_{i=1}^k\alpha_i\right)-\alpha_j^2\left(\sum_{i=1}^k\alpha_i\right) -\alpha_j^2}{\left(\sum_{i=1}^k\alpha_i+1\right)\left(\sum_{i=1}^k\alpha_i\right)^2} \\
                    \var(\theta_j) &= \frac{\alpha_j\left(\sum_{i=1}^k\alpha_i\right)- \alpha_j^2}{\left(\sum_{i=1}^k\alpha_i+1\right)\left(\sum_{i=1}^k\alpha_i\right)^2}
                \end{align*}
                Finally, 
                \begin{align*}
                    \cov(\theta_j,\theta_i) &= \E(\theta_j\theta_i) - \E(\theta_j)\E(\theta_i) \\
                    \cov(\theta_j,\theta_i) &= \frac{\Gamma(1+\alpha_j)\Gamma(1+\alpha_i)\Gamma(\sum_{i=1}^k\alpha_i)}{\Gamma(\alpha_i)\Gamma(\alpha_j)\Gamma(2+\sum_{i=1}^k\alpha_i)} - \frac{\alpha_j\alpha_i}{\left(\sum_{i=1}^k\alpha_i\right)^2} \\
                    \cov(\theta_j,\theta_i) &= \frac{\alpha_j\alpha_i}{\left(\sum_{i=1}^k\alpha_i+1\right)\sum_{i=1}^k\alpha_i} - \frac{\alpha_j\alpha_i}{\left(\sum_{i=1}^k\alpha_i\right)^2} \\
                    \cov(\theta_j,\theta_i) &= \frac{-\alpha_j\alpha_i}{\left(\sum_{i=1}^k\alpha_i+1\right)\left(\sum_{i=1}^k\alpha_i\right)^2} \\
                \end{align*}
                Qualitatively, as any particular $\alpha_i$ grows with the rest held constant, the probability of high values for $\theta_i$ grows. When all the $\alpha_j$ grow together, the distribution becomes more concentrated around the center where all $\theta_j$ are close together, and when all the $\alpha_j$ shrink together, the distribution becomes clustered around high values for one of the $\theta_j$ with the rest close to $0$
            \item[ii)]
                Let, for each $1\leq j \leq k$, let \(|\{j;X_j=k\}|=n_j\). Then,
                \begin{itemize}
                    \item[a)]
                        \begin{align*}
                            P(\theta \st X_1,...,X_n) &\propto P(X_1,...,X_n \st \theta)P(\theta) \\
                            P(\theta \st X_1,...,X_n) &\propto \left(\prod_{i=1}^k\theta_i^{n_i}\right)\left(\prod_{i=1}^k\theta_i^{\alpha_i-1}\right) \\
                            P(\theta \st X_1,...,X_n) &\propto \prod_{i=1}^k\theta_i^{n_i+\alpha_i-1}\\
                        \end{align*}
                        So \(\theta \st X_1,...,X_n\) has a Dirichlet distribution with parameters \((n_1+\alpha_1,...,n_k+\alpha_k)\)
                    \item[b)]
                        \begin{align*}
                            P(X_{n+1}=j \st X_1,...,X_n) &= \idotsint_{\sum_{i=1}^k\theta_i=1}P(X_{n+1}=j \st X_1,...,X_n,\theta)P(\theta\st X_1,...,X_n)d\theta_1...d\theta_k \\
                            P(X_{n+1}=j \st X_1,...,X_n) &= \frac{\Gamma(\sum_{i=1}^k n_i +\alpha_i)}{\prod_{i=1}^k\Gamma(n_i+\alpha_i)}\idotsint_{\sum_{i=1}^k\theta_i=1}\theta_j \prod_{i=1}^k\theta_i^{n_i+\alpha_i-1}d\theta_1...d\theta_k \\
                            P(X_{n+1}=j \st X_1,...,X_n) &= \frac{n_j+\alpha_j}{\sum_{i=1}^kn_i+\alpha_i}
                        \end{align*}
                        So we can conclude that \(X_{n+1}=j \st X_1,...,X_n\) has a multinomial distribution with probability vector
                        \[p = \left( \frac{n_1+\alpha_1}{\sum_{i=1}^kn_i+\alpha_i},...,\frac{n_k+\alpha_k}{\sum_{i=1}^kn_i+\alpha_i}\right) \]
                \end{itemize}
        \end{itemize}
    \item[3.]
        If $X$ is Poisson with mean $\theta$, then \(P(X=x\st \theta)\propto \theta^x e^{-\theta}\), which makes a prior that is something of the sort \(P(\theta)\propto \theta^a e^{b\theta}\), for some hyperparameters $a$, $b$, a natural choice. This is satistified by the gamma distribution, which is the conjugate prior. If we parameterize it the usual way with \(P(\theta)\propto \theta^{\alpha-1} e^{-\beta\theta}\), we get the posterior 
        \begin{align*}
            P(\theta \st X) &\propto P(X \st \theta)P(\theta) \\
            P(\theta \st X) &\propto \theta^{\sum x_i} e^{-n\theta}\theta^{\alpha-1} e^{-\beta\theta} \\
            P(\theta \st X) &\propto \theta^{\sum x_i+\alpha-1} e^{-(\beta+n)\theta}\\
        \end{align*}
        Giving us a gamma posterior with parameters \((\sum x_i+\alpha,\beta+n)\). \\
        \vspace{5mm} \\
        We can get the Jeffreys prior from the Fisher information:
        \begin{align*}
            p(\theta) &\propto \sqrt{I(\theta)} \\
            p(\theta) &\propto \sqrt{-\E\left[\frac{d^2}{d\theta^2} \log(f(X\st \theta))\right]} \\
            p(\theta) &\propto \sqrt{-\E\left[\frac{d^2}{d\theta^2} x\log(\theta)-\theta+c\right]} \\
            p(\theta) &\propto \sqrt{-\E\left[\frac{d}{d\theta} \frac{x}{\theta}-1\right]} \\
            p(\theta) &\propto \sqrt{-\E\left[-\frac{x}{\theta^2}\right]} \\
            p(\theta) &\propto \sqrt{\frac{1}{\theta}} \\
            p(\theta) &\propto \theta^{-\frac{1}{2}} \\
        \end{align*}
        With the Jeffreys prior, we get the posteriors
        \begin{align*}
            P(\theta \st x) &\propto P(x \st \theta)P(\theta) \\
            P(\theta \st x) &\propto \theta^{x} e^{-\theta}\theta^{-\frac{1}{2}} \\
            P(\theta \st x) &\propto \theta^{x-\frac{1}{2}} e^{-\theta} \\
            P(\theta \st cx) &\propto \theta^{cx-\frac{1}{2}} e^{-\theta} \\
        \end{align*}
        Versus, for the 'scale invariant' prior
        \begin{align*}
            P(\theta \st x) &\propto P(x \st \theta)P(\theta) \\
            P(\theta \st x) &\propto \theta^{x} e^{-\theta}\theta^{-1} \\
            P(\theta \st x) &\propto \theta^{x-1} e^{-\theta} \\
            P(\theta \st cx) &\propto \theta^{cx-1} e^{-\theta} \\
        \end{align*}
        The posteriors are similar, but the Jeffreys prior has a mean that is a bit closer to the MLE mean, which probably makes it preferable as far as a noninformative prior. As $c$ grows, they both converge to the MLE.
    \item[4.]
        We can derive the Jefferies prior from the Fisher Information:
        \begin{align*}
            I(p)_{i,j} &= -\E\left[H\left(\log(f(X\st p))\right)_{i,j}\right] \\
            I(p)_{i,j} &= -\E\left[H\left(c + \sum_{i=1}^k x_i\log(p_i)\right)_{i,j}\right] \\
            I(p)_{i,j} &= \begin{cases} 
                                    -\E\left[-\frac{x_i}{p_i^2}\right] & \mbox{if } i=j \\ 
                                                                     0 & \mbox{if } i\neq j 
                               \end{cases}  \\
            I(p)_{i,j} &= \begin{cases} 
                               \frac{n}{p_i} & \mbox{if } i=j \\ 
                                       0 & \mbox{if } i\neq j 
                               \end{cases} 
        \end{align*}
        Which makes \(|I(p)|^{\frac{1}{2}} = n^{\frac{k}{2}}\prod_{i=1}^k p_i^{-\frac{1}{2}}\). We can thus conclude
        \[p(\theta) \propto \prod_{i=1}^k p_i^{-\frac{1}{2}} \]
        Making it a Dirichlet prior with each $\alpha$ set to $\frac{1}{2}$.
    \item[5.]
        \begin{itemize}
            \item[a)]
                If we let $X$ be the training data, $G_j$ be the testing data, consisting of gene $g_{l,j}$ at locus $l$ for individual $j$, and $P_j$ denote the population of training sample $j$, then previously our model was:
                \begin{align*}
                    P(P_j \st G_j, X) &\propto P(P_j)P(G_j \st P_j, X) \\
                    P(P_j \st G_j, X) &\propto P(P_j)\prod_{l=1}^{24} P(g_{l,j} \st P_j, X) \\
                \end{align*}
                Where $P(g_{l,j} \st P_j=k, X)$ was the portion of the training population $k$ which had gene $g_{l,j}$ at locus $l$. Now however, I implemented a new model, where
                \begin{align*}
                    P(P_j \st G_j, X, \alpha) &\propto P(P_j)P(G_j \st P_j, X, \alpha) \\
                    P(P_j \st G_j, X, \alpha) &\propto P(P_j)\prod_{l=1}^{24} P(g_{l,j} \st P_j, X, \alpha) \\
                \end{align*}
                Where $P(g_{l,j} \st P_j, X, \alpha)$ is the posterior prediction distribution based on all the genes at locus $l$ in population $P_j$ with a uniform $\alpha$ Dirichlet distribution (as in question 2.ii.b). If $W$ is the set of genes at locus $l$, and $n_{w,P_j}$ is the count from the training sample of gene $w$ at locus $l$ in population $P_j$, then this quantity is:
                \[ P(g_{l,j}=v \st P_j, X, \alpha) = \frac{\alpha + n_{v,P_j}}{\sum_{w\in W} \alpha +n_{w,P_j}} \]
            \item[b)]
                Trying out different values of $\alpha$, I got these empirical error rates
                \FloatBarrier
                \input{hw3/pr5_b.tex}
                \FloatBarrier
            \item[c)]
                I calculated the likelihood for a given alpha as the outcome which made our training data most likely. In other words:
                \begin{align*}
                    P(G \st X,\alpha) &= \prod_{j=1}^n P(G_j \st X,\alpha) \\ 
                    P(G \st X,\alpha) &= \prod_{j=1}^n \prod_{l=1}^{24} P(g_{l,j} \st X, \alpha) \\
                    P(G \st X,\alpha) &= \prod_{j=1}^n \prod_{l=1}^{24}\sum_{k=1}^4 P(P_j=k)P(g_{l,j} \st P_j=k, X, \alpha) \\
                \end{align*}
                I maximized $\alpha$ over this quantity, with \(P(g_{l,j} \st P_j=k, X, \alpha)\) defined as above.
            \item[d)]
                When I did this, I found the optimal parameter to be $\alpha=.363$, which had a log likelihood of $-13621$ and yielded an error rate of $.231$. This rate is only a bit higher than the best rate observed in part b of $.216$.
            \item[e)]
                Where A=\(\{0.025, 0.05, 0.1, 0.2, 0.4, 0.8, 1.6, 3.2, 6.4\}\), I implemented the method as such:
                \begin{align*}
                    P(P_j \st G_j, X) &\propto P(P_j)P(G_j \st P_j, X) \\
                    P(P_j \st G_j, X) &\propto \sum_{\alpha\in A} P(\alpha)P(P_j)P(G_j \st P_j, X, \alpha) \\
                    P(P_j \st G_j, X) &\propto \sum_{\alpha\in A} \frac{1}{|A|}P(P_j)\prod_{l=1}^{24} P(g_{l,j} \st P_j, X, \alpha) \\
                \end{align*}
                This time, I got an error rate of $.243$, which is higher than the best $\alpha$ by it self by a few percent. This isn't surprising though, we are averaging the probabilities over a few $\alpha$ which we know have higher error rates than the best $\alpha$ we've seen. Of course, it may be that this value is really closer to what $\alpha$ "should" be given the true variation in the data.

        \end{itemize}

\end{itemize}

>>>>

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

STAT 347 Homework 2, Part 1 \hfill Aaron Maurer
\vspace{2mm}
\hrule
\vspace{2mm}
\begin{itemize}
    \item[A1:]
        Filling out the table, I got:
        \FloatBarrier
        \input{hw2/pr1}
        \FloatBarrier
        Overall, the majority of the variation not accounted for by the mean can be explained by the Fac(T) space, even though the other variables can account for part of this total variation too. We can see this by noting that (if $S$ is the entire space)
        \[ SS_{T/1} > SS_{t/1} + SS_{R/1}+ SS_{(t:r)/(T+R+r)} \geq SS_{S/T} \]

        Additionally we notice the spaces defined by each of three variables are not orthogonal to each other; otherwise, the sum of squares in the space defined by each would be identical to the quotient space with the other two variables removed. Additionally, a significant amount of the covariation between each subspace and the Prothrombin level is due to the intersection of the three subspaces. This said, each quotient space still accounts for its own covariation with Prothombin level. Finally, there is variation unaccounted for by the joint span of the three subspaces due to each variable, as indicated by the variation of the quotient space of the interaction space and the span of the three variable spaces.
    \item[A2:]
        Let us say the number of rows is $R$ \\
        \begin{itemize}
            \item[(i)] 
                The subspace will have dimension $R$.
                Algebraically this can be expressed as \(\E(Y_{ij})=\alpha_i+\alpha_b\). 
                This might be used to model a variable assigned to pairs, where order doesn't matter and each member contributes an amount independently. For instance, it might be used to model the energy released by burning one ounce of one fuel and one ounce of another fuel.
            \item[(ii)] 
                The subspace will have dimension $R$ and be orthogonal to the subspace in the prior problem. Algebraically it can be expressed as \(\E(Y_{ij})=\alpha_i-\alpha_b\). This might be used to model the difference between the first item in a pair and the second. For instance, this was the appropriate model for the voltage difference between two metals in an electrolytic solution.
            \item[(iii)] This subspace will have dimension ${R \choose 2}$. Algebraically, it can be expressed as $\E(Y_{ij})=\alpha_{ij}$, where $\alpha_{ij}=\alpha_{ji}$. This might be used to model an effect which is unique to particular unordered pairs. For instance, one might use this model the energy released from the reaction of one ounce of one chemical mixed with one ounce of another chemical.
            \item[(iv)] This subspace will have dimension ${R \choose 2}$. Algebraically, it can be expressed as $\E(Y_{ij})=\alpha_{ij}$, where $\alpha_{ij}=-\alpha_{ji}$ when $i\neq j$. This might be used to model an effect which is unique for particular pairs, with the sign flipped depending on the order. For instance, this might be used to model the difference in score between the first player to move and the second player to move in a game of go (we'll say someone can play themselves). 
        \end{itemize}
\end{itemize}
